---
title: "Does station density move property crime?"
description: "A fixed-effects panel across simulated local government areas. Worked example."
date: 2026-02-01
categories: [R, fixest, panel-econometrics]
# No image set: Quarto uses the first rendered figure as the card preview.
---

<!--
This is a worked EXAMPLE that renders standalone on synthetic data, so the site
builds before any real project is added. Replace it with a real write-up, or
delete it. It demonstrates the fixed section order, the shared ggplot theme, and
the accent-against-grey figure pattern.
-->

## The question

Does the density of police stations across local government areas (LGAs) predict
property crime rates, once we account for everything about an area that does not
change over time? This example uses simulated data to demonstrate the analysis
structure; the real project would use the Victorian LGA crime panel.

## Data and cleaning

A balanced panel of 79 simulated LGAs observed annually from 2014 to 2023. In a
real version the work is mostly in the cleaning: harmonising LGA names across
sources, resolving a composite key of area by year, and dropping all-zero rows
before the join. Here the data is generated below so the page is self-contained.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
source(here::here("theme/theme_portfolio.R"))

set.seed(42)
area_types <- c("Inner metro", "Outer metro", "Regional", "Rural")

panel <- expand_grid(area = 1:79, year = 2014:2023) |>
  mutate(
    type = area_types[1 + (area %% 4)],
    station_density = rnorm(n(), 2, 0.5),
    # true effect only in outer-metro areas
    effect = if_else(type == "Outer metro", -0.9, -0.1),
    crime_rate = 60 + effect * station_density +
      rnorm(n(), 0, 3) + as.numeric(factor(area)) * 0.05
  )
```

## Method, and why this one

Two-way fixed effects: area effects absorb anything constant within an LGA
(baseline affluence, geography), and year effects absorb shocks common to all
areas in a year (state-wide policy). The estimate then rests on within-area
change over time, not on comparing one area to another.

Why this rather than pooled OLS? Pooled OLS would let stable differences between
areas leak into the coefficient. Fixed effects is the standard defence against
that time-invariant confounding.

> Two-way fixed effects absorbs anything constant within an LGA, so the estimate
> rests on within-area change over time - not on comparing Melton to Melbourne.

```{r}
#| label: model
#| message: false
library(fixest)

m <- feols(crime_rate ~ station_density | area + year, data = panel)

# effect estimated separately by area type, for the figure
by_type <- panel |>
  group_split(type) |>
  map_dfr(\(d) {
    fit <- feols(crime_rate ~ station_density | area + year, data = d)
    tibble(type = unique(d$type),
           estimate = coef(fit)[["station_density"]],
           se = se(fit)[["station_density"]])
  })
```

The pooled coefficient on station density is
`r round(coef(m)[["station_density"]], 2)`. Broken out by area type, the effect
concentrates in outer-metro areas and is negligible elsewhere.

## Findings

```{r}
#| label: fig-effects
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Fig 1. The association concentrates in outer-metro areas; other area types show no detectable effect. Synthetic data."
by_type |>
  mutate(highlight = type == "Outer metro") |>
  ggplot(aes(x = fct_reorder(type, estimate), y = estimate, fill = highlight)) +
  geom_col(width = 0.65) +
  geom_errorbar(aes(ymin = estimate - 1.96 * se,
                    ymax = estimate + 1.96 * se),
                width = 0.15, colour = "grey40") +
  scale_fill_highlight(guide = "none") +
  coord_flip() +
  labs(
    title = "Station density and property crime, by area type",
    subtitle = "Estimated within-area effect with 95% intervals",
    x = NULL, y = "Effect on crime rate"
  ) +
  theme_portfolio()
```

## Assumptions and limitations

This design assumes that, after removing area and year effects, station density
is not driven by an omitted time-varying factor. That assumption is doing a lot
of work: if areas add stations in response to rising crime, the coefficient is
biased regardless of the fixed effects. The standard errors here are classical;
a real analysis would cluster by area to allow for serial correlation within an
LGA. And because the data is simulated, none of these numbers say anything about
the world - they only demonstrate the pipeline.

## Reproducing this

The page is self-contained: it generates its own data, so it renders with only
`tidyverse` and `fixest` installed. A real write-up would link to the project
repo and its `renv.lock`.
